Write-and-f-array: implementation and an application

### 2013-08-31 msc.tex
ujednolicone \var i \arrayspec

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{tikz}
\usepackage{comment}
\usetikzlibrary{arrows,decorations.pathreplacing}
\newcommand{\atomic}[4]{\node(lab#1) at (#3,-#4+0.5) {#2};\draw (#3,-#4-0.1) to (lab#1);\node[inner sep=0pt](#1) at (#3,-#4) {};}
\algblockdefx{Struct}{EndStruct}[1]{\textbf{struct} \textsc{#1}}{}
\algnewcommand{\algorithmicgoto}{\textbf{go to}}
\algnewcommand{\Goto}{\textbf{goto} }
\algnewcommand{\Label}{\State\unskip}
\def\none{\textbf{none} }
\def\T{\ensuremath{\operatorname{\mathcal{T}}}\text{ }}
\def\int{\ensuremath{\operatorname{\textbf{int}}}}
\algnotext{EndStruct}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\fn}[1]{\textsc{#1}}
\usepackage{float}
\newfloat{Listing}{t}{lop}
\author{Robert Obryk}
\title{Write-and-f-array: implementation and an application}
\date{}
\begin{document}

@@

\begin{algorithmic}[1]
\State \T $x$

@@

\begin{algorithmic}[1]
\State \textbf{int} $x$

@@

\begin{algorithmic}[1]
\State \textbf{int} $V$ \Comment{A CAS object holding an \textbf{int}}

@@

\subsection{Wait-freeness}

@@

\begin{algorithmic}[1]
\State \textbf{int} $T[P]$ \Comment{T is an array of $P$ CAS objects of type int}
\Function{inc$_p$}{}
	\State $x \gets T[p].\fn{rd}()$
	\State $T[p].\fn{wr}(x + 1)$
\EndFunction
\Function{read}{}
	\State $R \gets 0$
	\For{$i \gets 1 \mbox{to} P$}
		\State $R \gets R + T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
\State \T $T[N]$
\Function{update$_i$}{$x$}
	\State $T[i] \gets x$
\EndFunction
\Function{scan}{}
	\For{$i \gets 0 \mbox{to} N$}
		\State $R[i] \gets T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
\State \T $T[N]$
\Function{update-and-snap$_i$}{$x$}
	\State $T[i] \gets x$
	\For{$i \gets 0 \mbox{to} N$}
		\State $R[i] \gets T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
\State \T $T[N]$
\Function{update$_i$}{$x$}
	\State $T[i] \gets x$
\EndFunction
\Function{scan}{}
	\State\Return $f(T[0], T[1], \ldots, T[N-1])$
\EndFunction
\end{algorithmic}

@@

\begin{Listing}
\begin{algorithmic}[1]
\State \textbf{int} V
\Function{fetch-and-add}{$x$}
	\State $r \gets V$
	\State $V \gets V + x$
	\State\Return $r$
\EndFunction
\end{algorithmic}
\caption{Specification of semantics of fetch-and-add}
\label{faa-spec}
\end{Listing}

@@

\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State $\T H[]$
	\State $\int ver$
	\Function{get-current}{}
		\State \Return $ver$, $H[ver]$
	\EndFunction
	\Function{get}{$v$}
		\If{$v \leq ver - N$}
			\State \Return \none
		\Else
		\State \Return $H[ver]$
		\EndIf
	\EndFunction
	\Function{publish}{$v$, $T$}
		\If{$v = ver + 1$}
		\State $H[v] \gets T$
			\State $ver \gets v$
			\State \Return true
		\Else
			\State \Return false
		\EndIf
	\EndFunction
\end{algorithmic}
\caption{Specification of the history object}
\label{hist-spec}
\end{Listing}
We will first implement a helper object -- the history object. Intuitively, it contains a versioned memory cell and allows retrieval of past $N$ values of the cell. The cell holds objects of type \T.
Our wait-free implementation will impose an additional constraint on its use: every call to \fn{publish} is parametrized by an integer in range $[0,P)$ and executions of \fn{publish$_p$} for the same $p$ can not run simultaneously.
	Both $N$ and $P$ are known when the object is created. The exact semantics of the structure are given by the single-threaded specification in Listing~\ref{hist-spec}

@@

\begin{algorithmic}[1]
	\State $\textbf{pair}\left<\int v, \int p\right> S$ \Comment{Version number and $p$ of the most recently published value}
	\State $\textbf{pair}\left<\int v, \T T\right> H[N]$ \Comment{A circular buffer of last $N$ values (except maybe the most recent one)}
	\State $\textbf{pair}\left<\int v, \T T\right> L[P]$ \Comment{Temporary storage for values being published}
	\Function{get-current}{}
		\State $s \gets S$ \Comment{Possible serialization point}
		\State \fn{help}()
		\State \Return $H[s.v \mod N]$
	\EndFunction
	\Function{help}{}
		\State $s \gets S$
		\State $l \gets L[s.p]$
		\State $h \gets H[s.v \mod N]$
		\If{$l.v = s.v \textbf{and} h.v < s.v$}
		\State $\text{CAS}(H[s.v \mod N], h, l)$ \label{hist-help-cas}
		\EndIf
	\EndFunction
	\Function{get}{v}
		\State $s \gets S$ \label{hist-get-notyet-sp} \Comment{Possible serialization point}
		\If{$s.v < v$}
		\State \Return \none \label{hist-get-notyet} \Comment{The requested version hasn't been published yet}
		\EndIf
		\State \fn{help}()
		\State $h \gets H[v \mod N]$
		\If{$h.v = v$}
			\State \Return $h.T$
		\Else
			\State \Return \none \label{hist-get-tooold}
		\EndIf
	\EndFunction
	\Function{publish$_p$}{v, T}
		\State $s \gets S$ \label{hist-pub-earlier-sp} \Comment{Possible serialization point}
		\If{$v \neq s.v + 1$}
			\State \Return \textbf{false} \label{hist-pub-earlier-exit}
		\EndIf
		\State \fn{help}()
		\State $L[p] \gets (v, T)$ \label{hist-pub-latest}
		\If{$\text{CAS}(S, s, (v, p)$) \textbf{failed}} \Comment{Possible serialization point}\label{hist-pub-later-sp}
			\State \Return \textbf{false}
		\EndIf
		\State \Return \textbf{true}
	\EndFunction
\end{algorithmic}

@@

\begin{Listing}
\begin{algorithmic}[1]
	\State \T $v[N]$
	\State \textbf{int} $\text{last\_update}[N]$
	\State \textbf{int} $\text{last\_version}[N]$
	\State \textbf{T} $\text{last\_val}[N]$

@@

\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State \textbf{pair<int v, \T T>} S
	\Function{write-and-f$_0$}{T'}
		\State v, T $\gets$ S
		\State S $\gets$ (v + 1, T')
		\State \Return v + 1, v + 1, T
	\EndFunction
	\Function{get-last}{$i$}
		\State v, T $\gets$ S
		\State \Return v, v, T
	\EndFunction
	\Function{read}{}
		\State \Return S
	\EndFunction
\end{algorithmic}
\caption{The enriched write-and-f-array snapshot implementation for $N=1$}
\label{impl1}
\end{Listing}
\FloatBarrier

The implementation for $N > 1$ is presented below (interspersed with comments):
\begin{algorithmic}[1]
	\Struct{node-value}
		\State \T $T[2]$ \Comment{Child values}
		\State \textbf{int} $v[2]$ \Comment{Child versions}
	\EndStruct
	\Struct{last-value}
		\State \textbf{int} $n$
		\State \textbf{int} $v$
		\State \T $T$
	\EndStruct
	\State \textsc{write-and-f-array} $C[2]$ \Comment{of size resp. $\lceil{}N/2\rceil{}$ and $\lfloor{}N/2\rfloor{}$}
	\State \textbf{history object<node-value>} $H$ \Comment{of size $N+1$, for $N$ concurrent updates}
	\State \textbf{last-value} $L[N]$ %\Comment{thread's update count, version, value from before update}
	\algstore{impl}
\end{algorithmic}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{placeins}
\usepackage{amsthm}
\usepackage[noend]{algpseudocode}
\usepackage{tikz}
\usepackage{comment}
\usetikzlibrary{arrows,decorations.pathreplacing}
\newcommand{\atomic}[4]{\node(lab#1) at (#3,-#4+0.5) {#2};\draw (#3,-#4-0.1) to (lab#1);\node[inner sep=0pt](#1) at (#3,-#4) {};}
\algblockdefx{Struct}{EndStruct}[1]{\textbf{struct} \textsc{#1}}{}
\algnewcommand{\algorithmicgoto}{\textbf{go to}}
\algnewcommand{\Goto}{\textbf{goto} }
\algnewcommand{\Label}{\State\unskip}
\def\none{\textbf{none} }
\def\T{\ensuremath{\operatorname{\mathcal{T}}}\text{ }}
\def\int{\ensuremath{\operatorname{\textbf{int}}}}
\algnotext{EndStruct}
\newtheorem{lemma}{Lemma}
\newtheorem{observation}{Observation}
\newtheorem{theorem}{Theorem}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newcommand{\code}[1]{\text{#1}}
\newcommand{\fn}[1]{\textsc{#1}}
\newcommand{\var}[2]{\textbf{var }#1 : #2}
\newcommand{\arrayspec}[1]{\text{array}[#1]\text{ of }}
\usepackage{float}
\newfloat{Listing}{t}{lop}
\author{Robert Obryk}
\title{Write-and-f-array: implementation and an application}
\date{}
\begin{document}

@@

\begin{algorithmic}[1]
\State\var{$x$}{\T}

@@

\begin{algorithmic}[1]
\State\var{$x$}{\int}

@@

\begin{algorithmic}[1]
\State\var{$V$}{\int} \Comment{A CAS object holding an \int}

@@

\subsection{Wait-freedom}

@@

\begin{algorithmic}[1]
	\State\var{$T$}{\arrayspec{P}$\int$} \Comment{T is an array of $P$ CAS objects of type \int}
\Function{inc$_p$}{}
	\State $x \gets T[p].\fn{rd}()$
	\State $T[p].\fn{wr}(x + 1)$
\EndFunction
\Function{read}{}
	\State $R \gets 0$
	\For{$i \gets 1 \mbox{to} P$}
		\State $R \gets R + T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
	\State\var{$T$}{\arrayspec{N}$\T$}
\Function{update$_i$}{$x$}
	\State $T[i] \gets x$
\EndFunction
\Function{scan}{}
	\For{$i \gets 0 \mbox{to} N$}
		\State $R[i] \gets T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
	\State\var{$T$}{\arrayspec{N}$\T$}
\Function{update-and-snap$_i$}{$x$}
	\State $T[i] \gets x$
	\For{$i \gets 0 \mbox{to} N$}
		\State $R[i] \gets T[i]$
	\EndFor
	\State\Return $R$
\EndFunction
\end{algorithmic}

@@

\begin{algorithmic}[1]
	\State\var{$T$}{\arrayspec{N}$\T$}
\Function{update$_i$}{$x$}
	\State $T[i] \gets x$
\EndFunction
\Function{scan}{}
	\State\Return $f(T[0], T[1], \ldots, T[N-1])$
\EndFunction
\end{algorithmic}

@@

\begin{Listing}
\begin{algorithmic}[1]
\State\var{V}{\int}
\Function{fetch-and-add}{$x$}
	\State $r \gets V$
	\State $V \gets V + x$
	\State\Return $r$
\EndFunction
\end{algorithmic}
\caption{Specification of semantics of fetch-and-add}
\label{faa-spec}
\end{Listing}

@@

\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State\var{$H$}{\arrayspec{\ldots}$\T$}
	\State\var{ver}{\int} % fixme formatting
	\Function{get-current}{}
		\State \Return $ver$, $H[ver]$
	\EndFunction
	\Function{get}{$v$}
		\If{$v \leq ver - N$}
			\State \Return \none
		\Else
		\State \Return $H[ver]$
		\EndIf
	\EndFunction
	\Function{publish}{$v$, $T$}
		\If{$v = ver + 1$}
		\State $H[v] \gets T$
			\State $ver \gets v$
			\State \Return true
		\Else
			\State \Return false
		\EndIf
	\EndFunction
\end{algorithmic}
\caption{Specification of the history object}
\label{hist-spec}
\end{Listing}
We will first implement a helper object -- the history object. Intuitively, it contains a versioned memory cell and allows retrieval of past $N$ values of the cell. The cell holds objects of type \T.
Our wait-free implementation will impose an additional constraint on its use: every call to \fn{publish} is parametrized by an integer in range $[0,P)$ and executions of \fn{publish$_p$} for the same $p$ can not run simultaneously.
	Both $N$ and $P$ are known when the object is created. The exact semantics of the structure are given by the single-threaded specification in Listing~\ref{hist-spec}

@@

\begin{algorithmic}[1]
	\State\var{$S$}{$\textbf{pair}\left<\int v, \int p\right>$} \Comment{Version number and $p$ of the most recently published value}
	\State\var{$H$}{\arrayspec{N}$\textbf{pair}\left<\int v, \T T\right>$} \Comment{A circular buffer of last $N$ values (except maybe the most recent one)}
	\State\var{$L$}{\arrayspec{P}$\textbf{pair}\left<\int v, \T T\right>$} \Comment{Temporary storage for values being published}
	\Function{get-current}{}
		\State $s \gets S$ \Comment{Possible serialization point}
		\State \fn{help}()
		\State \Return $H[s.v \mod N]$
	\EndFunction
	\Function{help}{}
		\State $s \gets S$
		\State $l \gets L[s.p]$
		\State $h \gets H[s.v \mod N]$
		\If{$l.v = s.v \textbf{and} h.v < s.v$}
		\State $\text{CAS}(H[s.v \mod N], h, l)$ \label{hist-help-cas}
		\EndIf
	\EndFunction
	\Function{get}{v}
		\State $s \gets S$ \label{hist-get-notyet-sp} \Comment{Possible serialization point}
		\If{$s.v < v$}
		\State \Return \none \label{hist-get-notyet} \Comment{The requested version hasn't been published yet}
		\EndIf
		\State \fn{help}()
		\State $h \gets H[v \mod N]$
		\If{$h.v = v$}
			\State \Return $h.T$
		\Else
			\State \Return \none \label{hist-get-tooold}
		\EndIf
	\EndFunction
	\Function{publish$_p$}{v, T}
		\State $s \gets S$ \label{hist-pub-earlier-sp} \Comment{Possible serialization point}
		\If{$v \neq s.v + 1$}
			\State \Return \textbf{false} \label{hist-pub-earlier-exit}
		\EndIf
		\State \fn{help}()
		\State $L[p] \gets (v, T)$ \label{hist-pub-latest}
		\If{$\text{CAS}(S, s, (v, p)$) \textbf{failed}} \Comment{Possible serialization point}\label{hist-pub-later-sp}
			\State \Return \textbf{false}
		\EndIf
		\State \Return \textbf{true}
	\EndFunction
\end{algorithmic}

@@

\begin{Listing}
\begin{algorithmic}[1]
	\State\var{$v[N]$}{\arrayspec{N}$\T$}
	\State\var{$\text{last\_update}$}{\arrayspec{N}\int}
	\State\var{$\text{last\_version}$}{\arrayspec{N}\int}
	\State\var{$\text{last\_val}$}{\arrayspec{N}\T}

@@

\begin{Listing}[!h]
\begin{algorithmic}[1]
	\State\var{S}{\textbf{pair<int v, \T T>}}
	\Function{write-and-f$_0$}{T'}
		\State v, T $\gets$ S
		\State S $\gets$ (v + 1, T')
		\State \Return v + 1, v + 1, T
	\EndFunction
	\Function{get-last}{$i$}
		\State v, T $\gets$ S
		\State \Return v, v, T
	\EndFunction
	\Function{read}{}
		\State \Return S
	\EndFunction
\end{algorithmic}
\caption{The enriched write-and-f-array snapshot implementation for $N=1$}
\label{impl1}
\end{Listing}
\FloatBarrier

The implementation for $N > 1$ is presented below (interspersed with comments):
\begin{algorithmic}[1]
	\Struct{node-value}
		\State \var{$T$}{\arrayspec{2}\T} \Comment{Child values}
		\State \var{$v$}{\arrayspec{2}\int} \Comment{Child versions}
	\EndStruct
	\Struct{last-value}
		\State \var{$n$}{\int}
		\State \var{$v$}{\int}
		\State \var{$T$}{\T}
	\EndStruct
	\State \var{$C$}{\arrayspec{2}\textsc{write-and-f-array}}  \Comment{of size resp. $\lceil{}N/2\rceil{}$ and $\lfloor{}N/2\rfloor{}$}
	\State \var{$H$}{\textbf{history object<node-value>}} \Comment{of size $N+1$, for $N$ concurrent updates}
	\State \var{$L$}{\arrayspec{N}\textbf{last-value}} %\Comment{thread's update count, version, value from before update}
	\algstore{impl}
\end{algorithmic}
